---
title: "psych.251.reproducibility project - processing data file"
author: "kris evans"
date: "10/22/2020"
output: html_document
---
#### Packages ####
```{r}


library(data.table)
library(plyr)
library(ggplot2)
library(multilevel)
library(lme4)
library(lmSupport)
library(AICcmodavg)
library(pbkrtest)
library(boot)
library(haven)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#setwd
setwd("~/R Mac working folder")

#https://midus.colectica.org// general data repository good to see all the time points data and variables

load("/Users/krise/R Mac working folder/sommerfeldt2019/ICPSR_29282/DS0001/29282-0001-Data.rda") #bio data
load("/Users/krise/R Mac working folder/sommerfeldt2019/ICPSR_04652/DS0001/04652-0001-Data.rda") #M2 survey 
load("/Users/krise/R Mac working folder/sommerfeldt2019/ICPSR_22840/DS0001/22840-0001-Data-REST.rda") #Milwaukee 
load("/Users/krise/R Mac working folder/sommerfeldt2019/ICPSR_02760/DS0001/02760-0001-Data.rda") #M1 


#upload and save data  - data prep github - https://github.com/sashasomms/coherence_behavioral/blob/master/Prep_Coherence_MIDUSII_20180727.R
#original repo - https://github.com/sashasomms/coherence_behavioral/

#Project 4 MIDUS 2 - Biomarkers data RAW data upload and save as framework - https://www.icpsr.umich.edu/web/ICPSR/studies%20/29282#
#downloaded straight to R using .rda file and will output #need to confirm this is right data set as there are other data files here (DS0001 and DS0002) bsed on the original data upload looks like it is the DS001 data, we downloaded the R version of the data .rda


bio.d <- da29282.0001


 #Project 1- (MIDUS 2) - Look s like this is MIDUS2 part 1, not MIDUS 1 Survey-   RAW data upload and save as framework - https://www.icpsr.umich.edu/web/ICPSR/studies%20/04652#
#raw data sets offer multiple versions will take most similar path of primary data set as opposed to all of them but may need to return to retrieve them ##likely the MIDUS 04-06 midus 2

sur.d <- da04652.0001



#Project 1- (MIDUS 2)  Milwaukee African American Sample - need special permission - https://www.icpsr.umich.edu/web/NACDA/studies/22840

mil.d <- da22840.0001

#saveoutput data to easily retrieve


#MIDUS 1 p1 # Zygosity data from Midus I https://midus.colectica.org/
# Downloaded from colectica with variables: M2ID, ZYGCAT, TOT_SIBS  --not sure if downloaded whole data set or just these variables, looks like M2ID and TOT_SIBS are in the M1 sturvey but ZYGCAT is twin survey data, a different cohort.

m1.d <- da02760.0001


```


```{r}
#load data from personal dir
#### Read in Data #### helpful to view variables here too bio- https://midus.colectica.org/data/int.example/791f12ac-b75b-4ab2-a07c-887d20ca35ca/int.example/11f11291-99c9-4861-9008-eafe34544fc1/int.example/b595aa35-77cc-47b7-89a4-ed1dece9ec6b
# Project 4 (biomarker)

dfP4 = bio.d ##read in new data here, make similar to old processing names for ease #biomarker data M2P4
dfP1 = sur.d ##4652
dfP1M = mil.d #milwaukee data 22840
dfPZ = m1.d


names(dfP4)
summary(dfP4$SAMPLMAJ) #matches the data check from the codebook 29282
length(dfP4$M2ID) #matches the data check 1255



names(dfP1)
summary(dfP1$SAMPLMAJ) #matches the codebook data
length(dfP1$M2ID) #matches codebook data 4652

summary(dfP1M$BACRSEX) #matches the milwaukee data set 22840
length(dfP1M$M2ID) #matches, there is no SAMPLMAJ var


length(dfPZ$M2ID) #matches codebook for the M1 data 2760
```

##################################################
#### Clean & organize Project 1 (Survey) data ####
##################################################

```{r}

## MIDUS2 version PWB (7-items per sub-scale) -

PWB2O = c('B1SPWBA2', 'B1SPWBE2', 'B1SPWBG2', 'B1SPWBR2', 'B1SPWBU2', 'B1SPWBS2') # old names
PWB2 = c('autonomy2', 'envMast2', 'persGrow2', 'posRela2', 'purpLife2', 'selfAcce2') # new names

COPEO = c('B1SEMCOP', 'B1SPRCOP', 'B1SDENIA', 'B1SVENT', 'B1SDISEN', 'B1SREINT', 'B1SACTIV', 'B1SPLAN') # old names
COPE = c('COPEem', 'COPEprob', 'COPE_denial', 'COPE_vent', 'COPE_disengage', 'COPE_posReGrow', 'COPE_active', 'COPE_plan') # new names

## change 98s to NA - not sure why changed the 98s to NA either - note this time around idid not change 98s of cope to NA
#dfP1[PWB2O][dfP1[PWB2O] == 98] = NA
#dfP1[COPEO][dfP1[COPEO] == 98] = NA

# Look-see
varDescribe(dfP1[PWB2O])
varDescribe(dfP1[COPEO])


# Rename columns 
setnames(dfP1, old=PWB2O, new=PWB2)
setnames(dfP1, old=COPEO, new=COPE)
length(dfP1$M2ID)

```

```{r}

# Look-see
varDescribe(dfP1[PWB2])
varDescribe(dfP1[COPE])
sapply(dfP1[PWB2], class) # factors
sapply(dfP1[COPE], class) # factors

```

```{r}

# Convert to numeric 
dfP1[PWB2] = sapply(dfP1[PWB2], as.numeric)
dfP1[COPE] = sapply(dfP1[COPE], as.numeric)

```

```{r}

## Composite PWB
dfP1$pwb2 = varScore(dfP1, Forward = PWB2, MaxMiss = .0)
varDescribe(dfP1$pwb2)
```

```{r}
# Miscellaneous
P1miscO = c('B1PBYEAR','B1PRSEX','B1PF7A', 'B1PF2A', 'pwb2')
P1misc = c('birth_year','P1_sex','P1_race', 'P1_ethnicity', 'pwb2')
setnames(dfP1, old=P1miscO, new=P1misc)
```

```{r}
P1cols = c("M2ID", P1misc, PWB2, COPE)
# Subset P1 data
dfP1ss = dfP1[P1cols]
```

########################

############################################################
#### Clean & organize Project 1 (Survey) MILWAUKEE data ####
############################################################
```{r}

## MIDUS2 version PWB (7-items per sub-scale)
PWB2O = c('BASPWBA2', 'BASPWBE2', 'BASPWBG2', 'BASPWBR2', 'BASPWBU2', 'BASPWBS2') # old names
PWB2 = c('autonomy2', 'envMast2', 'persGrow2', 'posRela2', 'purpLife2', 'selfAcce2') # new names

COPEO = c('BASEMCOP', 'BASPRCOP', 'BASDENIA', 'BASVENT', 'BASDISEN', 'BASREINT', 'BASACTIV', 'BASPLAN') # old names
COPE = c('COPEem', 'COPEprob', 'COPE_denial', 'COPE_vent', 'COPE_disengage', 'COPE_posReGrow', 'COPE_active', 'COPE_plan') # new names

# Rename columns
setnames(dfP1M, old=PWB2O, new=PWB2)
setnames(dfP1M, old=COPEO, new=COPE)
length(dfP1M$M2ID)
```

```{r}

# Look-see
varDescribe(dfP1M[PWB2])
varDescribe(dfP1M[COPE])
sapply(dfP1M[PWB2], class) # numeric
sapply(dfP1M[COPE], class) # numeric
```

```{r}

## Calculate composite PWB
dfP1M$pwb2 = varScore(dfP1M, Forward = PWB2, MaxMiss = .0)
```

```{r}
# Miscellaneous
P1MmiscO = c( 'BACBYR', 'BACRSEX', 'BACF7A', 'BACF2A', 'pwb2')
P1Mmisc = c('birth_year', 'P1_sex', 'P1_race', 'P1_ethnicity', 'pwb2')
setnames(dfP1M, old=P1MmiscO, new=P1Mmisc)

class(dfP1$birth_year)
dfP1M$birth_year = sapply(dfP1M$birth_year, as.factor)


P1Mcols = c("M2ID", P1Mmisc, PWB2, COPE)
# Subset P1 data
dfP1Mss = dfP1M[P1Mcols]

```

```{r}

varDescribe(dfP1Mss)
varDescribe(dfP1ss)
# make sure no IDs are duplicated
summary(duplicated(c(dfP1ss$M2ID, dfP1Mss$M2ID)))

```

```{r}

####################################################
#### Clean & organize Project 4 (Biomarker) data ####
#####################################################
## Self reports of stress -- this is stress rating, baseline 
stressSRO = c('B4VSRB1', 'B4VSRCS1', 'B4VSRR1', 'B4VSRCS2', 'B4VSRR2', 'B4VSRU1') # old names --
stressSR = c('stress1', 'stress2', 'stress3', 'stress4', 'stress5', 'stress6') # new names baseline, cognitive stressor 1, recovery rating, cognitive stressor 2, recovery rating, stand/up right stressor


## Heart rate- B1 Heart rate, beats per minute, M1 Heart rate, R1 Heart rate, S1 Heart rate, R2 Heart rate, U1 HR, Avg heart rate,
#note there is baseline 2 measurement variable that isnt included could be useful B4VB2HR
#baseline, math, recovery, stroop, rest, up
ecg_HRO = c('B4VB1HR', 'B4VM1HR', 'B4VR1HR', 'B4VS1HR', 'B4VR2HR', 'B4VU1HR') # old names
ecg_HR = c('hr1', 'hrM', 'hr3', 'hrS', 'hr5', 'hr6') # new names


## ECG quality 
# Translate ECG quality ratings to number format 
old = c('(1) EXCELLENT', '(2) GOOD', '(3) SCOREABLE', '(4) UNSCOREABLE', '(5) NO DATA')
new = c(1, 2, 3, 4, 5)
dfP4$B4VBEQn = varRecode(dfP4$B4VBEQ, old, new)  #name variable from codebook - baseline ecg data quality 
dfP4$B4VMEQn = varRecode(dfP4$B4VMEQ, old, new) #math task ecg data quality
dfP4$B4VPEQn = varRecode(dfP4$B4VPEQ, old, new) #Pasat Math ECG Data Quality
dfP4$B4VSEQn = varRecode(dfP4$B4VSEQ, old, new) #Stroop Task ECG Data Quality
dfP4$B4VR1EQn = varRecode(dfP4$B4VR1EQ, old, new) #Recovery 1 ECG Data Quality
dfP4$B4VR2EQn = varRecode(dfP4$B4VR2EQ, old, new) #Recovery 2 ECG Data Quality
dfP4$B4VUEQn = varRecode(dfP4$B4VUEQ, old, new) #Standup Task ECG Data Quality



# c('B4VBEQ', 'B4VMEQ', 'B4VPEQ', 'B4VSEQ', 'B4VR1EQ', 'B4VR2EQ', 'B4VUEQ') # original names 
ecg_QO = c('B4VBEQn', 'B4VMEQn', 'B4VR1EQn', 'B4VSEQn', 'B4VR2EQn', 'B4VUEQn') # original names, in number format
ecg_Q = c('ecgQ1', 'ecgQM', 'ecgQ3', 'ecgQS', 'ecgQ5', 'ecgQ6') # new names baseline, math, pasat, recovery 1, recovery 2, stand up

## Miscellaneous variables to rename
#miscO = c('B1PGENDER'-DNE, 'B1PAGE_M2', 'B4ZAGE', 'B4ZB1PLG', 'B4ZB1SLG', 'B4ZB1CLG'-DNE THERE IS NO LAGP1 COG VARIABLE IN DATASET WILL ADJUST TO FIT B4ZB3CLG, 'B4QTA_AX', 'B4QCESD', 'B4H1I', 'B4PBMI', 'B4BIL6', 'B4BCRP')
#d[miscO]

#variables## sex, age determined by subtracting DOB, age during clinic, lag P1 phone to p4 complete in months, lag p1 SAW to p4, lag P3 cognitvive batter to P4, spielberge trait anxiety inventory,CESD: Center for Epidemiologic Studies Depression Scale, Ever had diabetes, BMI (Body Mass Index), Blood Serum IL6 (pg/mL), Blood C-Reactive Protein (ug/mL)


miscO = c('B1PRSEX', 'B1PAGE_M2', 'B4ZAGE', 'B4ZB1PLG', 'B4ZB1SLG', 'B4ZB3CLG', 'B4QTA_AX', 'B4QCESD', 'B4H1I', 'B4PBMI', 'B4BIL6', 'B4BCRP')
misc = c('gender', 'P1_PIage', 'P4_age', 'months_P1PI_to_P4', 'months_P1SAQ_to_P4', 'months_P1cog_to_P4', 'P4_STAItrait', 'P4_CESD', 'P4_diabetes', 'P4_BMI', 'IL6', 'CRP')


## Remove whitespace from task variable values - there is no B4VTASK1 var, B4VCS1 and 2 are the psychological test variables where subjects mark the test completed. - no need to recode but will rename to keep according
#dfP4$B4VTASK1str = varRecode(dfP4$B4VTASK1, c("STROOP    ", "MATH      ", "INAPPLIC  ", "PASAT     "), c('STROOP', 'MATH', 'INAPPLIC', 'PASAT'))


dfP4$B4VTASK1str = dfP4$B4VCS1

dfP4$B4VTASK2str = dfP4$B4VCS2


## Rename variables with my intuitive names

setnames(dfP4, old=ecg_HRO, new=ecg_HR)
setnames(dfP4, old=ecg_QO, new=ecg_Q)
setnames(dfP4, old=miscO, new=misc)
setnames(dfP4, old=stressSRO, new=stressSR) #add

```

```{r}

# Check some quality ratings before removing any bad data
varDescribeBy(dfP4$hr1, dfP4$ecgQ1) # 1 = 742, 2 = 364, 3 = 47 #mathes below
varDescribeBy(dfP4$hrM, dfP4$ecgQM) # 1 = 667, 2 = 363, 3 = 66
varDescribeBy(dfP4$hr3, dfP4$ecgQ3) # 1 = 718, 2 = 365, 3 = 42
varDescribeBy(dfP4$hrS, dfP4$ecgQS) # 1 = 663, 2 = 374, 3 = 76
varDescribeBy(dfP4$hr5, dfP4$ecgQ5) # 1 = 698, 2 = 378, 3 = 43
varDescribeBy(dfP4$hr6, dfP4$ecgQ6) # 1 = 716, 2 = 345, 3 = 43 

count(is.na(dfP4$hr1)) # 102 #match
count(is.na(dfP4$hrM)) # 159
count(is.na(dfP4$hr3)) # 130
count(is.na(dfP4$hrS)) # 142
count(is.na(dfP4$hr5)) # 136
count(is.na(dfP4$hr6)) # 151


```

```{r personal version}

dfP4 %>% count(is.na(hr1)) # confirm the above quality checks on the HR measure variable and data due to call issue
dfP4 %>% count(is.na(hrM))
dfP4 %>% count(is.na(hr3))
dfP4 %>% count(is.na(hrS))
dfP4 %>% count(is.na(hr5))
dfP4 %>% count(is.na(hr6))



dfP4$stress1
dfP4$stress2
dfP4$ecgQM

dfP4$B4VTASK1str
dfP4$hr1
dfP4$ecgQ1
```



```{r}
#### Clean ECG Data ####

## Replace all ecg data that is not good (2) or excellent (1) with NA 
dfP4['hr1'][ (dfP4['ecgQ1'] != 1) & (dfP4['ecgQ1'] != 2) ] = NA
dfP4['hrM'][ (dfP4['ecgQM'] != 1) & (dfP4['ecgQM'] != 2) ] = NA
dfP4['hr3'][ (dfP4['ecgQ3'] != 1) & (dfP4['ecgQ3'] != 2) ] = NA
dfP4['hrS'][ (dfP4['ecgQS'] != 1) & (dfP4['ecgQS'] != 2) ] = NA
dfP4['hr5'][ (dfP4['ecgQ5'] != 1) & (dfP4['ecgQ5'] != 2) ] = NA
dfP4['hr6'][ (dfP4['ecgQ6'] != 1) & (dfP4['ecgQ6'] != 2) ] = NA



# How many removed? 
count(is.na(dfP4$hr1)) # 149 - 102 = 47 #confirmed
count(is.na(dfP4$hrM)) # 225 - 159 = 66
count(is.na(dfP4$hr3)) # 172 - 130 = 42
count(is.na(dfP4$hrS)) # 218 - 142 = 76
count(is.na(dfP4$hr5)) # 179 - 136 = 43
count(is.na(dfP4$hr6)) # 194 - 151 = 43


```



```{r}

#### Clean diabetes data ####
dfP4$P4_diabetes = varRecode(dfP4$P4_diabetes, c("(1) YES", "(2) NO", "(3) BORDERLINE"), c(3, 1, 2)) # 1 = not diabetic, 2 = borderline diabetic, 3 = diabetic
dfP4$P4_diabetes = as.numeric(dfP4$P4_diabetes)
varDescribeBy(dfP4$M2ID, dfP4$P4_diabetes)
```


```{r}

#### Score BMI data ####
# Underweight: BMI is less than 18.5
# Normal weight: BMI is 18.5 to 24.9
# Overweight: BMI is 25 to 29.9
# Obese: BMI is 30 or more
# via: https://www.cancer.org/cancer/cancer-causes/diet-physical-activity/body-weight-and-cancer-risk/adult-bmi.html
varDescribe(dfP4$P4_BMI)

# Exclude underweight - why exclude underweight but not overweight? or just keep

dfP4['P4_BMI'][ (dfP4['P4_BMI'] < 18.5 ) ] = NA

```


```{r}

# Subset variables I want into a separate dataframe (dfP4ss) remove - 'B4VTASK1str' but kept psych test var "B4VCS1"
P4cols = c("M2ID", "M2FAMNUM", "SAMPLMAJ", 'B4VTASK1str', 'B4VCS1', 'B4VTASK2str', 'B4VCS2', misc, stressSR, ecg_HR, ecg_Q)
names(dfP4[P4cols])
dfP4ss = dfP4[P4cols]


```
```{r}
dfP4ss$P4_diabetes = as.numeric(dfP4$P4_diabetes) #1=yes, 2=no
```

```{r}
#### Sort out order of stressor tasks ####
# ECG variables are tied to the task they were collected during/after, but tasks were counterbalanced. 
# Stress self-reports are tied to the order they were measured.
# I need heart rate in order measured, task irrelevant, so can look at relation to stress self-report.  
summary(dfP4ss$B4VCS1)

## Subset stroop-firsts and maths-firsts into 2 separate dataframes
# dS = stroop first this doesnt return the proper data frame because 0 length
dSS = subset(dfP4ss, B4VTASK1str == 'Stroop')
length(dSS$M2ID) # 603
```

#####################################
#####################################
#####################################
#####################################
#####################################
#####################################
#####################################
#####################################
#####################################
#####################################
###### Skip to this line as this often causes issues adding in and switchin to tidyversere here
########
#####################################
#####################################
#####################################
#####################################
#####################################
#####################################
#########
#####################################

```{r}
##redo similarly because above call drops the stroop task try again
library(tidyverse)
tidyverse_conflicts()


dfP4ss$B4VTASK1str = as.numeric(dfP4$B4VTASK1str)
dfP4ss$B4VTASK2str = as.numeric(dfP4$B4VTASK2str)
ds <- dfP4ss %>% filter(B4VTASK1str == 2)

length(ds$M2ID) #602 - missing one participant at this point but not sure which one in comparison 
```

```{r}

# dM = math first - also does not workso will re do
dMo = subset(dfP4ss, B4VTASK1str == 'MATH')
length(dMo$M2ID) # 592
```

```{r}
dM <- dfP4ss %>% filter(B4VTASK1str == 1)
 #we do get the same participant data count at 592 here
```


```{r}

## Stroop first (stroop = 2, math = 4)
names(ds)[names(ds) == 'hrS'] = 'hr2'
names(ds)[names(ds) == 'hrM'] = 'hr4'
names(ds)[names(ds) == 'ecgQS'] = 'ecgQ2'
names(ds)[names(ds) == 'ecgQM'] = 'ecgQ4'

# Check did not edit other data frame like data.table sometimes does
names(ds)
names(dM) # good, still M and S

## Math first (math = 2, stroop = 4)
names(dM)[names(dM) == 'hrM'] = 'hr2'
names(dM)[names(dM) == 'hrS'] = 'hr4'
names(dM)[names(dM) == 'ecgQM'] = 'ecgQ2'
names(dM)[names(dM) == 'ecgQS'] = 'ecgQ4'
```


```{r}

# Merge stroop-first and math-first dataframes back together
dfP4ss2 = merge.data.frame(dM, ds, all=TRUE)

# New ECG variable names with numbers instead of 's' or 'm'
ecg_Qn = c('ecgQ1', 'ecgQ2', 'ecgQ3', 'ecgQ4', 'ecgQ5', 'ecgQ6')
ecg_HRn = c('hr1', 'hr2', 'hr3', 'hr4', 'hr5', 'hr6')

```


```{r}
#attempt to reproduce figure 1 which are histograms of self-report stress and HR over 5 time points for participants #match general display of data, need to adjust aesthetic

#baseline SR stress
ggplot(dfP4ss2, aes(x=stress1)) + 
  geom_histogram(color="black", fill="lightblue") +
  labs(title = "Baseline SR Stress Histogram", x="Self-reported Stress", y="Number of Participants") +
theme_bw()

#Stressor task 1
ggplot(dfP4ss2, aes(x=stress2)) + 
  geom_histogram(color="black", fill="lightblue") +
labs(title = "Stressor Task 1 (Stroop/Math) Stress Histogram", x="Self-reported Stress", y="Number of Participants")

#recovery 1
ggplot(dfP4ss2, aes(x=stress3)) + 
  geom_histogram(color="black", fill="lightblue") +
labs(title = "Recovery 1 Stress Histogram", x="Self-reported Stress", y="Number of Participants")


#stressor task 2
ggplot(dfP4ss2, aes(x=stress4)) + 
  geom_histogram(color="black", fill="lightblue") +
labs(title = "Stressor Task 2 (Stroop/Math) Stress Histogram", x="Self-reported Stress", y="Number of Participants")

#Recovery 2
ggplot(dfP4ss2, aes(x=stress5)) + 
  geom_histogram(color="black", fill="lightblue") +
labs(title = "Recovery 2 Stress Histogram", x="Self-reported Stress", y="Number of Participants")
```

```{r}
#summary statistics - match figure 1

stress1.summary <- summary(dfP4ss2$stress1, na.rm = TRUE)

stress2.summary <- summary(dfP4ss2$stress2, na.rm = TRUE)

stress3.summary <- summary(dfP4ss2$stress3, na.rm = TRUE)

stress4.summary <- summary(dfP4ss2$stress4, na.rm = TRUE)

stress5.summary <- summary(dfP4ss2$stress5, na.rm = TRUE)


stress1.summary
stress2.summary
stress3.summary
stress4.summary
stress5.summary
sd(dfP4ss2$stress1, na.rm = TRUE)
sd(dfP4ss2$stress2, na.rm = TRUE)
sd(dfP4ss2$stress3, na.rm = TRUE)
sd(dfP4ss2$stress4, na.rm = TRUE)
sd(dfP4ss2$stress5, na.rm = TRUE)

```


```{r}
#attempt to reproduce figure 1 which are histograms of HR over 5 time points for participants #match general display of data, need to adjust aesthetic

#baseline SR stress
ggplot(dfP4ss2, aes(x=hr1)) + 
  geom_histogram(color="black", fill="lightblue") +
  labs(title = "Baseline SR Stress Histogram", x="Heart Rate (Beats per minute)", y="Number of Participants") +
theme_bw()

#Stressor task 1
ggplot(dfP4ss2, aes(x=hr2)) + 
  geom_histogram(color="black", fill="lightblue") +
labs(title = "Stressor Task 1 (Stroop/Math) Stress Histogram", x="Heart Rate (Beats per minute)", y="Number of Participants")

#recovery 1
ggplot(dfP4ss2, aes(x=hr3)) + 
  geom_histogram(color="black", fill="lightblue") +
labs(title = "Recovery 1 Stress Histogram", x="Heart Rate (Beats per minute)", y="Number of Participants")


#stressor task 2
ggplot(dfP4ss2, aes(x=hr4)) + 
  geom_histogram(color="black", fill="lightblue") +
labs(title = "Stressor Task 2 (Stroop/Math) Stress Histogram", x="Heart Rate (Beats per minute)", y="Number of Participants")

#Recovery 2
ggplot(dfP4ss2, aes(x=hr5)) + 
  geom_histogram(color="black", fill="lightblue") +
labs(title = "Recovery 2 Stress Histogram", x="Heart Rate (Beats per minute)", y="Number of Participants")



#summary statistics - match figure 1

hr1.summary <- summary(dfP4ss2$hr1, na.rm = TRUE)
hr2.summary <- summary(dfP4ss2$hr2, na.rm = TRUE)
hr3.summary <- summary(dfP4ss2$hr3, na.rm = TRUE)
hr4.summary <- summary(dfP4ss2$hr4, na.rm = TRUE)
hr5.summary <- summary(dfP4ss2$hr5, na.rm = TRUE)


hr1.summary
hr2.summary
hr3.summary
hr4.summary
hr5.summary

sd(dfP4ss2$hr1, na.rm = TRUE)
sd(dfP4ss2$hr2, na.rm = TRUE)
sd(dfP4ss2$hr3, na.rm = TRUE)
sd(dfP4ss2$hr4, na.rm = TRUE)
sd(dfP4ss2$hr5, na.rm = TRUE)
```


##################### Due to versions and packages should complete the data wrangling for bio data below, followed by final merge
###################### Merge datasets to final frameworks
##################

```{r}

#### Merge P1 and P1 Milwaukee ####
dfP1_P1M = rbind(dfP1ss, dfP1Mss)
varDescribe(dfP1_P1M)
########################
```

```{r}

#subset data frame - taking zygosity variables - # Zygosity data from Midus I 
# Downloaded from colectica with variables: M2ID, ZYGCAT, TOT_SIBS
dfPZ2 <- dfPZ %>% dplyr::select(M2ID, ZYGCAT, TOT_SIBS)

#####################
####  Merge data #### 
#####################
# Merge zyogsity and twins data to P1 data
dfTemp2 = merge.data.frame(dfP1_P1M, dfPZ2,  by='M2ID', all=TRUE)
dfTemp3 = merge.data.frame(dfP1_P1M, dfPZ2, by='M2ID', all = FALSE)
```

```{r final data mere}

# Merge that zygosity/twins & P1 data to P4 data
dfTemp = merge.data.frame(dfTemp2, dfP4ss2,  by='M2ID', all=TRUE)


df = dfTemp

```


```{r}

#saveoutput data framework to easily retrieve upload and manipulate
write.csv(df,"psych.251.repro.df.csv")

#can upload this data frame above to begin from here

```

```{r}
#keep rows that have stroop or math task as the first task and remove NAs
df1 <- df %>% filter(B4VTASK1str == 1 | B4VTASK1str == 2, na.rm = TRUE)

#where are the NAS?
summary(is.na(df1))

#drop additional NAs from the second task , drops 30 more
df2 <- df1 %>% filter(B4VTASK2str == 1 | B4VTASK2str == 2, na.rm = TRUE)


summary(is.na(df1))
```


```{r}

################################
### Convert to long format  #### 
################################
## Long format 
# Make a copy of the dataframe to convert to a data table 
# because data tables link to all versions of table, which causes issues when trying to change column names in only one copy
dt = df

# All variables except stress[1-6], hr[1-6], ecgQ[1-6]
names(dt)
varsid = c(names(dt[1:37]), names(dt[56:57]))
dfLTemp = melt.data.table(setDT(dt),
                       id.vars = varsid, # ID variables - all the variables to keep but not split apart on
                       measure = patterns("^stress", "^hr", "^ecgQ"),
                       variable.name = "timepoint")
OLD = c("value1", "value2", "value3")
NEW = c("stress", "hr", "ecgQ")
setnames(dfLTemp, old=OLD, new=NEW)
```

```{r}

# Convert to data frame
dfLTemp = data.frame(dfLTemp)
names(dfLTemp)

```


```{r}
#### Remove orthostatic stress timepoint from long format df ####
dfLnoO = dfLTemp[dfLTemp$timepoint != 6,] 

dfL = dfLnoO

```

##################################
##################################
##################################
##################################
##################################
##################################skip here too 
##################################
######have to do all of the above and then use this data frame to continue on as has all the pertinent variables
##################################
##################################
##################################
##################################
##################################
##################################
##################################
```{r}

#saveoutput data framework to easily retrieve upload and manipulate
#can either  manipulate and work from the dt this was right before data made long at 687
write.csv(dt,"psych.251.repro.dt.csv")
```

```{r}

#saveoutput data framework to easily retrieve upload and manipulate
#can either  manipulate and work from the dfP4ss2 data set (which was merged a little above) or from the df, which is the merged data frame and should still have the variables used in the chunks below. IF work on this data set, be sure to go through merge updates again, can also move that chunk following the bio data cleaning
write.csv(dfP4ss2,"psych.251.repro.dfP4ss2.csv")


```
